---
title: 'MA304 : An Exploratory Data Visualisation Study of Weather-Related Crime Trends in Colchester (2024)'
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
author: "Presented by : 2400589_Sakthivel Vinayagam"
---

## Introduction

Understanding **crime dynamics** is essential for effective **public safety strategies**. In recent years, researchers have increasingly explored how external factors, such as _weather conditions_, may influence criminal activity. This report examines the role of _climatic variables_ on **crime rates** in **Colchester, Essex**, during the year **2024**.

### Overview of the Datasets

This investigation uses two datasets: **crime24.csv**, which includes monthly _street-level crime incidents_ in **Colchester** (e.g., crime type, location, outcome), and **temp24.csv**, which provides daily _weather data_ such as **temperature**, **humidity**, **pressure**, and **precipitation** from a nearby station.

### Aim and Objectives

The analysis begins by independently exploring both **crime** and **climate** datasets to identify _seasonal_, _temporal_, and _geographic patterns_. These datasets are then merged to examine how fluctuations in **temperature**, **humidity**, or **pressure** align with changes in _crime volume_ or _type_. A variety of **visualisation techniques**, including _time series_, _scatter plots_, and _spatial mapping_, are applied to highlight key patterns and associations. The ultimate goal is to derive practical, **actionable recommendations** for local law enforcement based on _data-informed insights_.

### Research Question

> **Are crime levels influenced by weather in Colchester?**

To answer this, we’ll perform:
- Exploratory data analysis and visualisation of both datasets
- Time series analysis and smoothing
- Mapping and spatial insight using Leaflet
- Correlation and clustering techniques to explore links between climate and crime

Through this investigation, we aim to uncover trends that may assist in **strategic policing, crime prevention**, and **public awareness** campaigns tailored to seasonal or climatic patterns.

------------------------------------------------------------------------

## Data Preparation

```{r}

### Load required packages
library(dplyr)
library(ggplot2)
library(lubridate)
library(readr)
library(forcats)
library(leaflet)
library(plotly)
library(tidyr)
library(reshape2)
library(patchwork)
library(viridis)
library(gganimate)

```

### Load the datasets

```{r}
# Load the crime dataset from CSV file
crime24 <- read_csv("crime24.csv", show_col_types = FALSE)

# Display the first few rows of the dataset
as.data.frame(head(crime24))
```

```{r}
# Load the temperature dataset from CSV files
temp24 <- read_csv("temp24.csv", show_col_types = FALSE)

# Display the first few rows of the dataset
as.data.frame(head(temp24))
```

### Convert date format and extract useful time features

```{r}
# Fix incomplete date format by adding a day to form valid dates (e.g., "2024-01" -> "01-2024-01")
crime24$date <- paste0("01-", crime24$date)  
crime24$date <- as.Date(crime24$date, format = "%d-%Y-%m")

# Extract full month name from the date
crime24$month <- month(crime24$date, label = TRUE, abbr = FALSE)

# Assign a season based on the month number
crime24$season <- case_when(
  month(crime24$date) %in% c(12, 1, 2) ~ "Winter",   # December–February
  month(crime24$date) %in% c(3, 4, 5) ~ "Spring",    # March–May
  month(crime24$date) %in% c(6, 7, 8) ~ "Summer",    # June–August
  month(crime24$date) %in% c(9, 10, 11) ~ "Autumn"   # September–November
)

# Remove unused columns with no analytical value
crime24 <- crime24 %>% select(-c(context, location_subtype))

# Replace missing values in outcome_status with "Unknown"
crime24$outcome_status <- ifelse(is.na(crime24$outcome_status), "Unknown", crime24$outcome_status)

# Display a count of missing values by column
colSums(is.na(crime24))
```

### Clean and prepare temperature dataset

```{r}
# Convert the 'Date' column to proper Date format
temp24$Date <- as.Date(temp24$Date)

# Extract the full month name from the date
temp24$month <- month(temp24$Date, label = TRUE, abbr = FALSE)

# Assign a season based on the month number
temp24$season <- case_when(
  month(temp24$Date) %in% c(12, 1, 2) ~ "Winter",   # December–February
  month(temp24$Date) %in% c(3, 4, 5) ~ "Spring",    # March–May
  month(temp24$Date) %in% c(6, 7, 8) ~ "Summer",    # June–August
  month(temp24$Date) %in% c(9, 10, 11) ~ "Autumn"   # September–November
)

# Remove unnecessary columns with excessive missing values
temp24 <- temp24 %>% select(-c(PreselevHp, SnowDepcm))

# Replace missing values in rainfall and low cloud cover columns with 0
temp24$Precmm[is.na(temp24$Precmm)] <- 0
temp24$lowClOct[is.na(temp24$lowClOct)] <- 0

# Display a count of remaining missing values by column
colSums(is.na(temp24))
```

## Exploratory Data Analysis (EDA)

### Crime Category Distribution

```{r}
# Create a bar plot showing the distribution of crime categories, ordered by frequency
crime_category_plot <- ggplot(crime24, aes(x = fct_infreq(category), fill = category)) +
  geom_bar(color = "black") +  # Add black borders to bars
  labs(title = "Distribution of Crime Categories",
       x = "Crime Category", y = "Frequency") +
  theme_minimal() +  # Apply a clean, minimal theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

# Convert the static ggplot to an interactive plotly version
ggplotly(crime_category_plot)
```

The interactive bar plot shows that **violent crime** is the most frequently reported offence in **Colchester**, followed by _anti-social behaviour_. Other crimes like _shoplifting_ and _criminal damage_ occur moderately, while **robbery** and **possession of weapons** are less common. Overall, _violence_ and _anti-social behaviour_ dominate the crime profile in **2024**.

### Outcome Status of Crimes

```{r}
# Create a bar plot showing the distribution of crime outcome statuses, ordered by frequency
outcome_plot <- ggplot(crime24, aes(x = fct_infreq(outcome_status), fill = outcome_status)) +
  geom_bar(color = "black") +  # Add black outline around bars
  labs(title = "Outcome Status of Reported Crimes",
       x = "Outcome Status", y = "Count") +
  theme_minimal() +  # Use a clean minimal theme
  theme(
    axis.text.x = element_text(angle = 70, hjust = 1),  # Tilt x-axis labels for better readability
    axis.title = element_text(size = 12),               # Set font size for axis titles
    plot.title = element_text(size = 16, face = "bold") # Make plot title larger and bold
  ) +
  scale_fill_manual(values = rainbow(length(unique(crime24$outcome_status)))) +  # Use a distinct color for each outcome
  guides(fill = guide_legend(title = "Outcome Status"))  # Add a legend title

# Convert the static plot to an interactive plotly version
ggplotly(outcome_plot)
```

The bar plot reveals that most reported crimes result in **no suspect being identified**, followed by cases where suspects _cannot be prosecuted_. A notable portion of cases remains _under investigation_ or _await court outcomes_, while fewer end in _formal charges_ or other resolutions. This suggests that many crimes in **Colchester** do not lead to prosecution, highlighting challenges in _investigation_ and _legal follow-through_.

### Crime by Location Type

```{r}
# Create a bar chart showing the number of crimes by location type
location_plot <- ggplot(crime24, aes(x = location_type, fill = location_type)) +
  geom_bar(color = "black") +  # Add black borders to each bar
  labs(title = "Crimes by Location Type",
       x = "Location Type", y = "Number of Crimes") +
  theme_minimal()  # Apply a clean, minimal visual theme

# Convert the ggplot chart into an interactive Plotly version
ggplotly(location_plot)
```

The bar chart highlights that nearly all crimes are reported under the local police authority (**“Force”**), with very few linked to the **British Transport Police** (_“BTP”_). This suggests that most criminal incidents occur in _community settings_ rather than _transport environments_, emphasizing the primary role of **local policing** in managing crime.

### Top 10 Crime Hotspot Streets

```{r}
# Identify the top 10 streets with the highest number of crimes
top_streets <- crime24 %>%
  group_by(street_name) %>%
  summarise(crime_count = n()) %>%
  arrange(desc(crime_count)) %>%
  slice_head(n = 10)

# Filter the original crime dataset to include only those top 10 streets
crime_top <- crime24 %>% filter(street_name %in% top_streets$street_name)

# Create a stacked bar plot showing crime type distribution across the top 10 streets
hotspot_plot <- ggplot(crime_top, aes(x = street_name, fill = category)) +
  geom_bar() +  # Plot counts by street and crime category
  labs(title = "Crime Type Distribution in Top 10 Streets",
       x = "Street", y = "Number of Crimes") +
  scale_fill_viridis_d(option = "plasma") +  # Apply a visually appealing color palette
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Tilt x-axis labels for readability

# Make the chart interactive with Plotly
ggplotly(hotspot_plot)
```

The bar plot shows that **“On or near Supermarket”** and **“On or near Shopping Area”** are the top crime hotspots, largely driven by _shoplifting_. In contrast, locations like _nightclubs_ and _police stations_ show a more diverse mix of crimes, including _violent offences_ and _public disorder_. This suggests that **crime patterns vary by location**, with _retail areas_ experiencing targeted offences and other zones showing broader criminal activity.

### Spatial Mapping of Crime (Leaflet Map)

```{r}
# Create an interactive leaflet map to display crime locations across Colchester
leaflet(crime24) %>%
  addTiles() %>%  # Add default OpenStreetMap tile layer as the base map
  addCircleMarkers(
    lng = ~long, lat = ~lat,       # Use longitude and latitude for marker placement
    radius = 3,                    # Set marker size
    color = "blue",                # Use blue for marker color
    stroke = FALSE,                # Remove border around circles
    fillOpacity = 0.6              # Set transparency for better visibility
  ) %>%
  setView(                         # Center the map view on the average location of all crimes
    lng = mean(crime24$long, na.rm = TRUE),
    lat = mean(crime24$lat, na.rm = TRUE),
    zoom = 13                      # Set zoom level for city-scale view
  )
```

The interactive **leaflet map** reveals that _central Colchester_, particularly around **major roads** and _commercial zones_, is the main crime hotspot. Areas like **Cymbeline Way** and the _town centre_ show the highest concentration of incidents, while _outer residential areas_ see far fewer. This indicates that crime is closely tied to _high-traffic_, _public-facing spaces_.

### Monthly Crime Trend Over Time

```{r}
# Ensure the 'date' column is in proper Date format
crime24$date <- as.Date(crime24$date)

# Create a new variable representing month and year (e.g., "2024-03")
crime24$month_year <- format(crime24$date, "%Y-%m")

# Group the dataset by month and count the number of crimes in each
monthly_crimes <- crime24 %>%
  group_by(month_year) %>%
  summarise(total_crimes = n()) %>%
  mutate(month_year = as.Date(paste0(month_year, "-01")))  # Convert to Date for plotting

# Generate a line plot showing monthly crime trends, with a loess smoothing line
monthly_trend_plot <- ggplot(monthly_crimes, aes(x = month_year, y = total_crimes)) +
  geom_line(color = "steelblue", size = 1) +  # Line for monthly crime totals
  geom_smooth(method = "loess", se = FALSE, color = "darkred", linetype = "dashed") +  # Smoothed trend line
  labs(title = "Monthly Crime Trend in Colchester (2024)",
       x = "Month", y = "Number of Crimes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

# Convert to an interactive Plotly plot
ggplotly(monthly_trend_plot)
```

The **line chart** shows that crime in **Colchester** peaked mid-year, with the highest levels in **July**, and declined toward the end of **2024**. Early months like _January_ and _February_ also saw elevated activity, while _April_ marked a low point. The overall trend suggests a _seasonal pattern_, with increased crime during **warmer months**, possibly due to higher _public activity_.

### Crime Count Across Seasons

```{r}
# Group crime data by season and count total crimes in each
seasonal_crimes <- crime24 %>%
  group_by(season) %>%
  summarise(total_crimes = n()) %>%
  arrange(desc(total_crimes))  # Order seasons by crime count (descending)

# Create a bar plot showing the number of crimes in each season
season_plot <- ggplot(seasonal_crimes, aes(x = season, y = total_crimes, fill = season)) +
  geom_bar(stat = "identity", color = "black") +  # Use black borders on bars
  scale_fill_manual(values = c("Winter" = "#4682B4", "Spring" = "#90EE90",
                               "Summer" = "#FFD700", "Autumn" = "#FF8C00")) +  # Assign custom colors
  labs(title = "Crime Count Across Seasons (2024)",
       x = "Season", y = "Number of Crimes") +
  theme_minimal()  # Apply a clean visual style

# Convert to an interactive plot using Plotly
ggplotly(season_plot)

```

The **bar chart** shows that **Summer** had the highest crime levels in **Colchester** during **2024**, followed by **Autumn**, while _Spring_ recorded the fewest crimes. This suggests that crime slightly increases during _warmer months_, likely due to greater _public activity_, and decreases in _colder seasons_. Overall, there is a modest _seasonal pattern_ in crime rates.

### Heatmap of Crime Categories Across Seasons

```{r}
# Count the number of crimes grouped by both season and category
crime_season_heatmap <- crime24 %>%
  group_by(season, category) %>%
  summarise(crime_count = n(), .groups = 'drop')  # Drop grouping after summarising

# Create a heatmap to show the distribution of crime types across seasons
heatmap_plot <- ggplot(crime_season_heatmap, aes(x = season, y = category, fill = crime_count)) +
  geom_tile(color = "white") +  # Draw rectangles with white borders
  scale_fill_gradient(low = "#ffffcc", high = "#006837", name = "Crime Count") +  # Color gradient
  labs(title = "Crime Types by Season in Colchester (2024)",
       x = "Season", y = "Crime Category") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),         # Rotate x-axis labels
    plot.title = element_text(size = 16, face = "bold"),       # Style the title
    legend.position = "right"                                  # Position the legend
  )

# Convert the heatmap to an interactive version
ggplotly(heatmap_plot)
```

The **heatmap** shows that **violent crime** is consistently the most reported offence across all seasons in **Colchester**, with slight peaks in _summer_ and _winter_. Other common crimes like _anti-social behaviour_ and _shoplifting_ occur steadily year-round, while less frequent crimes show little seasonal change. Overall, _crime patterns_ remain fairly stable throughout the year, with only modest _seasonal variation_.

## Weather Data Trends

### Time Series of Avg, Max & Min Temperature

```{r}
# Reshape temperature data into a tidy (long) format for plotting
temp_long <- temp24 %>%
  select(Date, TemperatureCAvg, TemperatureCMax, TemperatureCMin) %>%
  pivot_longer(
    cols = c(TemperatureCAvg, TemperatureCMax, TemperatureCMin),  # Select columns to reshape
    names_to = "Metric",      # New column for metric names (Avg, Max, Min)
    values_to = "Temperature" # New column for temperature values
  )

# Create a time series line plot of average, max, and min daily temperatures
temp_trend_plot <- ggplot(temp_long, aes(x = Date, y = Temperature, color = Metric)) +
  geom_line(size = 0.7) +  # Use thin lines for clarity
  labs(title = "Temperature Trends in Colchester (2024)",
       x = "Date", y = "Temperature (°C)", color = "Type") +
  scale_color_manual(values = c(  # Assign specific colors for each metric
    "TemperatureCAvg" = "blue",
    "TemperatureCMax" = "red",
    "TemperatureCMin" = "green"
  )) +
  theme_minimal()  # Apply a clean visual theme

# Convert the plot to an interactive version with Plotly
ggplotly(temp_trend_plot)
```

The **line chart** shows a clear _seasonal temperature cycle_ in **Colchester** during **2024**. Temperatures rise from _winter_ to _summer_, peaking in **July** and **August**, then gradually decline toward the year’s end. **Maximum**, _average_, and **minimum temperatures** follow consistent patterns, reflecting typical _UK climate trends_. This seasonal variation provides useful context for understanding how _weather may influence crime activity_.

### Temperature Distribution

```{r}
# Create a violin plot to show the distribution of average daily temperatures
temp_violin_plot <- ggplot(temp24, aes(y = TemperatureCAvg, x = "", fill = "")) +
  geom_violin(fill = "#87CEFA", color = "black") +  # Sky blue fill with black outline
  labs(
    title = "Distribution of Average Daily Temperature",
    y = "Average Temperature (°C)",
    x = ""
  ) +
  theme_minimal()  # Apply a clean, minimal visual theme

# Convert the static violin plot to an interactive version using Plotly
ggplotly(temp_violin_plot)
```

The **violin plot** shows that most _average daily temperatures_ in **Colchester** during **2024** fell between **7°C and 18°C**, with the highest concentration around **12–14°C**. _Extreme temperatures_ were rare, with few days below **5°C** or above **20°C**. Overall, this indicates a _mild and stable climate_ throughout the year.

### Monthly Temperature Summary

```{r}
# Calculate monthly average, maximum, and minimum temperatures
monthly_temp <- temp24 %>%
  mutate(month = month(Date, label = TRUE, abbr = FALSE)) %>%
  group_by(month) %>%
  summarise(
    AvgTemp = mean(TemperatureCAvg, na.rm = TRUE),
    MaxTemp = mean(TemperatureCMax, na.rm = TRUE),
    MinTemp = mean(TemperatureCMin, na.rm = TRUE)
  )

# Create a bar chart to show monthly average temperatures with error bars for range
monthly_temp_summary_plot <- ggplot(monthly_temp, aes(x = month)) +
  geom_bar(aes(y = AvgTemp), stat = "identity", fill = "#FFA07A") +  # Bar for average temperature
  geom_errorbar(aes(ymin = MinTemp, ymax = MaxTemp), width = 0.2) +  # Error bar for temperature range
  labs(
    title = "Monthly Average Temperature Range",
    x = "Month", y = "Temperature (°C)"
  ) +
  theme_minimal()  # Use a clean, minimal theme

# Convert the static plot to an interactive Plotly version
ggplotly(monthly_temp_summary_plot)
```

The **bar chart with error bars** shows a clear _seasonal temperature pattern_ in **Colchester** during **2024**. Temperatures rise from **January**, peak in **July** and **August**, then decline into _winter_. **Summer months** show both _higher averages_ and _wider temperature ranges_, while **winter months** are cooler and more stable. Overall, the plot highlights _typical seasonal shifts_ and _daily variability_ in temperature.

### Seasonal Summary

```{r}
# Create a boxplot to compare average daily temperatures across seasons
seasonal_temp_boxplot <- ggplot(temp24, aes(x = season, y = TemperatureCAvg, fill = season)) +
  geom_boxplot() +  # Display median, quartiles, and outliers
  scale_fill_manual(values = c(
    "Winter" = "#4682B4",   # Steel blue
    "Spring" = "#90EE90",   # Light green
    "Summer" = "#FFD700",   # Gold
    "Autumn" = "#FF8C00"    # Dark orange
  )) +
  labs(
    title = "Average Temperature by Season",
    x = "Season", y = "Average Temperature (°C)"
  ) +
  theme_minimal()  # Apply a clean, minimal style

# Convert the static boxplot to an interactive Plotly version
ggplotly(seasonal_temp_boxplot)
```

The **boxplot** highlights a clear _seasonal shift in temperatures_ in **Colchester** during **2024**. **Summer** is the warmest season, with a median above **17°C**, while **winter** is the coldest, with values near **7°C** and some days _below freezing_. _Spring_ and _autumn_ fall in between, with similar medians around **11–12°C**. These patterns align with the expected _seasonal temperature cycle_.

### Precipitation Density Plot

```{r}
# Create a density plot to show the distribution of daily precipitation values
precip_density_plot <- ggplot(temp24, aes(x = Precmm)) +
  geom_density(fill = "#6495ED", alpha = 0.7, color = "black") +  # Fill with cornflower blue and black border
  labs(
    title = "Density of Daily Precipitation",
    x = "Precipitation (mm)", y = "Density"
  ) +
  theme_minimal()  # Apply a clean minimal theme

# Convert the density plot to an interactive version using Plotly
ggplotly(precip_density_plot)
```


The **density plot** shows that most days in **Colchester** during **2024** were _dry_ or had _very light rain_, with a strong peak near **0 mm** of precipitation. **Heavier rainfall** (over **10 mm**) was rare, and _extreme rain events_ above **20 mm** were even less frequent. Overall, the year was marked by _mild_ and _mostly dry weather conditions_.

### Humidity & Pressure Time Series Plot

```{r}
# Reshape the data into long format for humidity and pressure
humidity_pressure <- temp24 %>%
  select(Date, HrAvg, PresslevHp) %>%
  pivot_longer(
    cols = c(HrAvg, PresslevHp),     # Columns to reshape
    names_to = "Variable",           # New column for variable names
    values_to = "Value"              # New column for values
  )

# Create a time series line plot of humidity and pressure over the year
humidity_pressure_ts_plot <- ggplot(humidity_pressure, aes(x = Date, y = Value, color = Variable)) +
  geom_line() +  # Add lines for each variable
  labs(
    title = "Humidity and Pressure Over Time",
    x = "Date", y = "Value", color = "Metric"
  ) +
  theme_minimal()  # Apply a minimal visual theme

# Convert the static plot to an interactive Plotly version
ggplotly(humidity_pressure_ts_plot)
```

The **line chart** shows that _humidity_ in **Colchester** remained fairly steady throughout **2024**, mostly between **80–90%**, with only slight fluctuations. **Atmospheric pressure** varied a bit more, ranging from **1000 to 1025 hPa**, with some shifts in _late winter_ and _spring_. Overall, both variables were _stable_, indicating _consistent atmospheric conditions_ during the year.

### Wet vs Dry Days

```{r}
# Classify each day as "Wet" if precipitation ≥ 1 mm, otherwise "Dry"
temp24$rain_type <- ifelse(temp24$Precmm >= 1, "Wet", "Dry")

# Create a bar plot comparing the number of wet and dry days
wet_dry_days_plot <- ggplot(temp24, aes(x = rain_type, fill = rain_type)) +
  geom_bar(color = "black") +  # Add black borders to bars
  scale_fill_manual(values = c("Dry" = "#FFD700", "Wet" = "#00BFFF")) +  # Gold for dry, blue for wet
  labs(
    title = "Count of Wet vs Dry Days in 2024",
    x = "Day Type", y = "Number of Days"
  ) +
  theme_minimal()  # Apply a clean minimal theme

# Convert the static plot to an interactive version using Plotly
ggplotly(wet_dry_days_plot)
```

The **bar chart** shows that **dry days** far outnumbered **wet days** in **Colchester** during **2024**, with more than _twice as many dry days_ recorded. This supports the earlier findings that the year was _mostly dry_, with rainfall occurring only _occasionally_, confirming a _mild_ and _dry climate pattern_ overall.

### Visibility Distribution

```{r}
# Create a density plot showing the distribution of daily visibility in kilometers
visibility_density_plot <- ggplot(temp24, aes(x = VisKm)) +
  geom_density(fill = "#6A5ACD", alpha = 0.7, color = "black") +  # Purple fill with black outline
  labs(
    title = "Density of Visibility (km)",
    x = "Visibility (km)", y = "Density"
  ) +
  theme_minimal()  # Apply a clean minimal theme

# Convert the static density plot to an interactive Plotly chart
ggplotly(visibility_density_plot)
```

The **visibility distribution** shows that in **2024**, most days in **Colchester** had _good visibility_, typically between **30–40 km**. _Low-visibility days_ were rare, indicating that _foggy_ or _hazy conditions_ were uncommon. Overall, the city experienced _clear_ and _stable atmospheric conditions_ throughout the year.

### Cloud Cover Distribution

```{r}
# Create a histogram showing the distribution of low cloud cover (in octas)
low_cloud_hist_plot <- ggplot(temp24, aes(x = lowClOct)) +
  geom_histogram(bins = 10, fill = "#708090", color = "black") +  # Slate gray fill with black borders
  labs(
    title = "Distribution of Low Cloud Cover (Octas)",
    x = "Low Cloud Cover (0–8)", y = "Frequency"
  ) +
  theme_minimal()  # Apply a clean, minimal theme

# Convert the static histogram to an interactive Plotly version
ggplotly(low_cloud_hist_plot)
```

The **histogram** shows that _low cloud cover_ in **Colchester** during **2024** was most frequently between **6 and 8 octas**, indicating _mostly cloudy_ to _overcast conditions_ on many days. _Clearer skies_ (low cloud cover below **4 octas**) were relatively rare. Overall, the city experienced _frequent cloudiness_ throughout the year.

### Wind Speed vs Gust Correlation

```{r}
# Create a scatter plot to show the relationship between wind speed and wind gusts
wind_speed_gust_plot <- ggplot(temp24, aes(x = WindkmhInt, y = WindkmhGust)) +
  geom_point(alpha = 0.5, color = "#20B2AA") +  # Semi-transparent teal points
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add linear trend line in red
  labs(
    title = "Correlation between Wind Speed and Wind Gust",
    x = "Wind Speed (km/h)", y = "Wind Gust (km/h)"
  ) +
  theme_minimal()  # Apply a clean, minimal theme

# Convert the scatter plot to an interactive Plotly version
ggplotly(wind_speed_gust_plot)
```

The **scatter plot** shows a strong _positive correlation_ between **wind speed** and **wind gusts** in **Colchester** during **2024**. As wind speed increases, wind gusts also tend to rise, following a fairly _linear pattern_. This suggests that on _windier days_, stronger gusts are more likely, which is typical in _weather dynamics_.

### Wind Direction Distribution

```{r}
# Prepare frequency data for wind directions
wind_dir_freq <- temp24 %>%
  count(WindkmhDir, name = "Count") %>%
  rename(Direction = WindkmhDir)  # Rename for clarity

# Define a custom color palette using the viridis scale
custom_colors <- viridis(length(unique(wind_dir_freq$Direction)))

# Create a polar bar chart (wind rose) using Plotly
plot_ly(wind_dir_freq,
        type = 'barpolar',                  # Use polar bar chart
        r = ~Count,                         # Radius represents count of days
        theta = ~Direction,                 # Angle represents wind direction
        color = ~Direction,                 # Use direction to define color groups
        colors = custom_colors) %>%
  layout(
    title = "Wind Direction Rose Chart (2024)",
    polar = list(
      angularaxis = list(
        direction = "clockwise",            # Rotate clockwise
        rotation = 90                       # Start from North
      ),
      radialaxis = list(
        showticklabels = FALSE,             # Hide tick labels for cleaner look
        ticks = ""                          # Remove tick marks
      )
    ),
    showlegend = FALSE                      # Hide legend for simplicity
  )
```

The **wind rose chart** shows that _westerly_ and _southwesterly winds_ were the most frequent in **Colchester** during **2024**, while _eastern_ and _northeastern winds_ were less common. This pattern reflects the _prevailing wind directions_ typical of the **UK’s maritime climate**, providing useful context for understanding _weather patterns_ and _environmental conditions_.

## Exploring Weather’s Impact on Crime Rates

### Merging crime24.csv & temp24.csv

```{r}
# Format dates for merging: extract "YYYY-MM" from each date
crime24$merge_date <- format(crime24$date, "%Y-%m")
temp24$merge_date <- format(temp24$Date, "%Y-%m")

# Aggregate weather data by month: calculate mean values for each variable
temp_monthly <- temp24 %>%
  group_by(merge_date) %>%
  summarise(
    avg_temp = mean(TemperatureCAvg, na.rm = TRUE),
    max_temp = mean(TemperatureCMax, na.rm = TRUE),
    min_temp = mean(TemperatureCMin, na.rm = TRUE),
    humidity = mean(HrAvg, na.rm = TRUE),
    pressure = mean(PresslevHp, na.rm = TRUE),
    precipitation = mean(Precmm, na.rm = TRUE)
  )

# Aggregate total number of crimes by month
crime_monthly <- crime24 %>%
  group_by(merge_date) %>%
  summarise(
    total_crimes = n()
  )

# Merge crime and weather datasets by month
weather_crime <- left_join(crime_monthly, temp_monthly, by = "merge_date")

# View the structure of the merged dataset
as.data.frame(head(weather_crime))
```



### Correlation Between Crime and Weather Variables
```{r}
# Remove the merge_date column and compute a correlation matrix for numeric variables
cor_matrix <- cor(weather_crime[, -1], use = "complete.obs")

# Reshape the correlation matrix into long format for plotting
cor_melted <- melt(cor_matrix)

# Round correlation values to two decimal places for labels
cor_melted$label <- round(cor_melted$value, 2)

# Create an interactive correlation heatmap using Plotly
crime_weather_corr_plot <- plot_ly(
  data = cor_melted,
  x = ~Var1,
  y = ~Var2,
  z = ~value,
  type = "heatmap",
  text = ~label,                      # Display correlation values as text
  texttemplate = "%{text}",           # Format text appearance
  textfont = list(color = "black", size = 12),
  hoverinfo = "text",                 # Show only text on hover
  colorscale = "YlGnBu",              # Apply Yellow-Green-Blue sequential color scale
  zmin = -1, zmax = 1,                # Set color scale limits
  colorbar = list(title = "Correlation")
) %>%
  layout(
    title = "Correlation Between Crime & Weather Variables",
    xaxis = list(title = "", tickangle = -45),
    yaxis = list(title = "")
  )


# Render the interactive correlation heatmap
crime_weather_corr_plot
```

The **correlation matrix** reveals that _crime in Colchester_ during **2024** is _moderately positively correlated_ with **temperature** and **precipitation**, suggesting higher crime rates during _warmer_ and _wetter conditions_. In contrast, **humidity** and **pressure** show _weak negative correlations_, indicating little to no influence on crime levels. Overall, **temperature** and **rainfall** appear to have the _strongest links_ to _crime trends_.


### Scatter Plots: Weather vs Crime

```{r}
# Plot 1: Relationship between average temperature and total crimes
p1 <- ggplot(weather_crime, aes(x = avg_temp, y = total_crimes)) +
  geom_point(color = "#E67E22", size = 3) +  # Orange points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Linear regression line
  labs(
    title = "Avg Temperature vs Crime",
    x = "Average Temperature (°C)",
    y = "Total Crimes"
  ) +
  theme_minimal()

# Plot 2: Relationship between average humidity and total crimes
p2 <- ggplot(weather_crime, aes(x = humidity, y = total_crimes)) +
  geom_point(color = "#3498DB", size = 3) +  # Blue points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Linear regression line
  labs(
    title = "Humidity vs Crime",
    x = "Average Humidity (%)",
    y = "Total Crimes"
  ) +
  theme_minimal()

# Plot 3: Relationship between average pressure and total crimes
p3 <- ggplot(weather_crime, aes(x = pressure, y = total_crimes)) +
  geom_point(color = "#2ECC71", size = 3) +  # Green points
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Linear regression line
  labs(
    title = "Pressure vs Crime",
    x = "Average Pressure (hPa)",
    y = "Total Crimes"
  ) +
  theme_minimal()

# Convert each ggplot object into interactive Plotly plots
p1_plot <- ggplotly(p1)
p2_plot <- ggplotly(p2)
p3_plot <- ggplotly(p3)

# Combine the three interactive plots into a single row layout
subplot(p1_plot, p2_plot, p3_plot, nrows = 1, margin = 0.05, titleX = TRUE, titleY = TRUE) %>%
  layout(
    title = list(
      text = "Weather Factors and Their Relationship with Crime",
      x = 0.5,              # Center the main title
      xanchor = "center",
      y = 0.95
    )
  )
```

The **scatter plots** show that _crime in Colchester_ tends to rise with **higher temperatures**, suggesting a _positive link_ between **warmth** and _criminal activity_. In contrast, **humidity** and **pressure** show slight _negative correlations_ with crime, though these relationships are weaker. Overall, **temperature** appears to be the _strongest weather-related factor_ associated with monthly crime patterns in **2024**.


### Crime on Wet vs Dry Months
```{r}
# Categorize each month as "Wet" or "Dry" based on average monthly precipitation
weather_crime$rain_type <- ifelse(weather_crime$precipitation >= 1, "Wet", "Dry")

# Create a boxplot comparing total crime counts between wet and dry months
wet_dry_crime_boxplot <- ggplot(weather_crime, aes(x = rain_type, y = total_crimes, fill = rain_type)) +
  geom_boxplot() +  # Display distribution, median, and range of crimes per group
  scale_fill_manual(values = c("Wet" = "#1F78B4", "Dry" = "#FFD700")) +  # Blue for wet, gold for dry
  labs(
    title = "Crime Comparison: Wet vs Dry Months",
    x = "Month Type", y = "Total Crimes"
  ) +
  theme_minimal()  # Apply a clean minimal theme

# Convert the static boxplot into an interactive Plotly version
ggplotly(wet_dry_crime_boxplot)
```

The **boxplot** compares _crime levels_ between **wet** and **dry months** in **Colchester** for **2024**. It shows that _wet months_ tend to have slightly _higher crime counts_ on average and a _wider range of variation_. This pattern suggests a _modest association_ between **rainfall** and _increased crime_, possibly due to more _social activity_ or _weather-related disruptions_ during wetter periods.


### Average Temperature Range vs Crime Types
```{r}
# Join crime and weather datasets by merge_date (month), allowing multiple matches
combined <- left_join(crime24, temp24, by = "merge_date", relationship = "many-to-many")

# Categorize temperature into three ranges using quantiles: Low, Medium, High
combined$temp_range <- cut(
  combined$TemperatureCAvg,
  breaks = quantile(combined$TemperatureCAvg, probs = c(0, 0.33, 0.66, 1), na.rm = TRUE),
  labels = c("Low", "Medium", "High"),
  include.lowest = TRUE
)

# Create a proportional bar plot showing crime category distribution by temperature range
crime_by_temp_range_plot <- ggplot(combined, aes(x = temp_range, fill = category)) +
  geom_bar(position = "fill") +  # Fill bars proportionally by category
  labs(
    title = "Crime Type Distribution by Temperature Range",
    x = "Temperature Range", y = "Proportion",
    fill = "Crime Category"
  ) +
  scale_fill_viridis_d(option = "magma") +  # Apply a perceptually uniform color palette
  theme_minimal()  # Use a clean visual theme

# Convert the static plot to an interactive Plotly version
ggplotly(crime_by_temp_range_plot)
```

The **chart** shows that while most _crime types_ occur consistently across _temperature ranges_, **anti-social behaviour** and **public order offences** are slightly more common in _warmer conditions_. In contrast, crimes like **vehicle crime** and **shoplifting** remain steady regardless of temperature. This suggests that some _crime types_ may be more influenced by **temperature** than others.

### Humidity Range vs Crime Types 
```{r}
# Categorize humidity into two bins: Low and High based on median split
combined$humidity_range <- cut(
  combined$HrAvg,
  breaks = quantile(combined$HrAvg, probs = c(0, 0.5, 1), na.rm = TRUE),
  labels = c("Low", "High"),
  include.lowest = TRUE
)

# Create a violin plot to visualize humidity distribution across crime categories
humidity_crime_violin_plot <- ggplot(combined, aes(x = category, y = HrAvg, fill = humidity_range)) +
  geom_violin(scale = "width", trim = FALSE, color = "black") +  # Violin plot with uniform width
  labs(
    title = "Distribution of Humidity by Crime Category",
    x = "Crime Category", y = "Humidity (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

# Convert the violin plot to an interactive Plotly chart
ggplotly(humidity_crime_violin_plot)
```

The **violin plot** shows that most _crime types_ occur under both _low_ and _high humidity_, but slightly more crimes—especially _violent_ and _public order offences_—are associated with **higher humidity levels**. This suggests a _minor trend_ of increased crime during more _humid conditions_.

## Visual Analytics-Driven Approaches to Advanced Clustering Techniques

### Interactive Leaflet Map with Clustering
```{r}
# Filter out rows with missing latitude or longitude to ensure valid map points
crime_map_data <- crime24 %>% filter(!is.na(lat), !is.na(long))

# Create an interactive leaflet map with marker clustering for crime incidents
leaflet(crime_map_data) %>%
  addTiles() %>%  # Add default OpenStreetMap tile layer
  addMarkers(
    lng = ~long, lat = ~lat,  # Use coordinates from dataset
    clusterOptions = markerClusterOptions(),  # Enable clustering of nearby markers
    popup = ~paste("Crime:", category, "<br>", "Street:", street_name)  # Display info on click
  ) %>%
  setView(
    lng = mean(crime_map_data$long, na.rm = TRUE),  # Center the map on the average longitude
    lat = mean(crime_map_data$lat, na.rm = TRUE),   # Center the map on the average latitude
    zoom = 13  # Set zoom level for detailed town view
  )
```

The **interactive leaflet map with clustering** provides a _spatial overview_ of **crime hotspots** across **Colchester** in **2024**. Each numbered circle represents a _cluster_ of reported crimes in a specific geographic area, with the number indicating the _total incidents_ in that location. _Larger circles_ and _warmer colors_ (orange/red) reflect areas with _higher concentrations of crime_.

From the map, it’s clear that the **town center**—particularly the vicinity around **Colchester Town Station** and **Southway**—emerges as a primary hotspot, with cluster counts exceeding **1,100 crimes** in some zones. Other significant concentrations appear along **Butt Road**, **Hythe Hill**, and near **St Nicholas Street**, suggesting these may be _high-traffic_ or _densely populated areas_. In contrast, peripheral areas such as _Cymbeline Meadows_ or _Abbey Field_ show much lower crime frequencies, with some clusters having fewer than **10 incidents**.

In summary, this _clustered map_ helps visualize how crime is _spatially distributed_, highlighting _urban centers_ as key hotspots. It enables authorities and policymakers to _target specific areas_ for surveillance, prevention, or community engagement efforts.

### K-Means Clustering of Crime Hotspots
```{r}
# Prepare data for K-means clustering
crime_k <- crime_map_data %>%
  select(lat, long) %>%
  na.omit()

# Run K-means clustering (4 clusters here, adjust as needed)
set.seed(123)
crime_kmeans <- kmeans(crime_k, centers = 4)

# Add cluster assignments to your spatial data
crime_map_data$cluster <- factor(crime_kmeans$cluster)
# Define a color palette for clusters
cluster_palette <- colorFactor(
  palette = c("red", "blue", "green", "purple"),
  domain = crime_map_data$cluster
)

# Create a Leaflet map showing crime clusters from K-means results
leaflet(crime_map_data) %>%
  addTiles() %>%  # Add default OpenStreetMap tiles
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    color = ~cluster_palette(cluster),  # Color markers by cluster
    radius = 5,
    stroke = FALSE,
    fillOpacity = 0.6,
    popup = ~paste("Cluster:", cluster, "<br>Crime:", category, "<br>Street:", street_name)
  ) %>%
  setView(
    lng = mean(crime_map_data$long, na.rm = TRUE),
    lat = mean(crime_map_data$lat, na.rm = TRUE),
    zoom = 14
  ) %>%
  addLegend("bottomright",
            pal = cluster_palette,
            values = ~cluster,
            title = "Crime Clusters",
            opacity = 0.8)
```


The **K-Means clustering plot** illustrates how _crime locations_ in **Colchester** are _spatially grouped_ based on geographic coordinates. Using four cluster centers, the algorithm identifies areas where incidents are _densely concentrated_, assigning each to a specific _color-coded cluster_.

From the visualization, we can distinguish four prominent _spatial clusters_. **Cluster 1** (_red_) is concentrated in the _southwestern zone_, **Cluster 2** (_blue_) covers the _eastern region_, **Cluster 3** (_green_) appears in the _southern-central part_, and **Cluster 4** (_purple_) is focused in the _northern stretch_ of the town center. This clear geographic separation suggests that _different neighborhoods_ in **Colchester** are subject to _distinct crime patterns_.

These insights are valuable for _law enforcement_ and _urban planners_. Cluster-specific trends could indicate hotspots for _nightlife_, _commercial activity_, or _residential vulnerability_. As a result, interventions—such as increased patrols, surveillance, or community outreach—can be _strategically localized_, allowing for more _effective and efficient crime prevention efforts_ across the town.


### Interactive Time Slider for Monthly Crime
```{r}
# Group crime data by month and category, and count the number of incidents
crime_time <- crime24 %>%
  group_by(month_year = format(date, "%Y-%m"), category) %>%
  summarise(count = n(), .groups = 'drop')

# Create an interactive multi-line chart showing monthly crime trends per category
plot_ly(
  data = crime_time,
  x = ~month_year,                 # X-axis: month and year
  y = ~count,                      # Y-axis: number of crimes
  color = ~category,               # Color lines by crime category
  colors = viridis_pal(option = "D")(length(unique(crime_time$category))),  # Apply perceptually uniform palette
  type = 'scatter', mode = 'lines+markers'  # Line plot with markers
) %>%
  layout(
    title = "Interactive Monthly Crime by Category",
    xaxis = list(title = "Month"),
    yaxis = list(title = "Crime Count")
  )
```

The **interactive chart** shows _monthly crime trends by category_ in **2024**. **Violent crime** is the most frequent throughout the year, followed by _anti-social behaviour_ and _shoplifting_. Some categories remain _low and stable_, such as _bicycle theft_ and _possession of weapons_. Overall, the chart highlights _seasonal fluctuations_ and helps identify which _crime types dominate each month_.


### Animated Time Series with gganimate
```{r}
# Prepare monthly crime data by summarising the number of incidents per category each month
crime_anim_data <- crime24 %>%
  mutate(month = format(date, "%Y-%m")) %>%
  group_by(month, category) %>%
  summarise(crime_count = n(), .groups = 'drop')

# Create a base ggplot object for animation
crime_anim_plot <- ggplot(crime_anim_data, aes(x = reorder(category, -crime_count), y = crime_count, fill = category)) +
  geom_col(show.legend = FALSE) +  # Use column chart without legend
  labs(
    title = 'Monthly Crime by Category: {closest_state}',  # Dynamic title showing current month
    x = 'Crime Category',
    y = 'Number of Crimes'
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Tilt x-axis labels for clarity
  transition_states(month, transition_length = 2, state_length = 1) +  # Animate across months
  ease_aes('cubic-in-out')  # Smooth easing for transitions

# Render the animation as a GIF (requires gifski package)
animate(crime_anim_plot, nframes = 100, fps = 10, renderer = gifski_renderer())

```

The **animated bar chart** visualizes how _crime categories_ changed _month-by-month_ throughout **2024**. It clearly shows **violent crime** consistently dominating across all months, followed by _anti-social behaviour_ and _shoplifting_. While the order of less frequent crimes shifts slightly, overall trends remain _stable_.

This animation provides a _dynamic view_ of _seasonal or monthly fluctuations_ in crime types, making it easier to spot _peaks_ and _patterns over time_. It’s especially useful for identifying _months with spikes_ in certain offences and tracking how _priorities may shift_ throughout the year.

### Animated Crime Hotspot Map (Over Time)
```{r}
# Prepare spatial crime data with valid coordinates and a monthly timestamp
crime_map_anim <- crime24 %>%
  filter(!is.na(lat) & !is.na(long)) %>%
  mutate(month = format(date, "%Y-%m"))

# Identify the top 5 most frequent crime categories
top_categories <- crime_map_anim %>%
  count(category, sort = TRUE) %>%
  top_n(5) %>%
  pull(category)

# Filter the dataset to include only the top 5 crime categories
crime_map_anim <- crime_map_anim %>%
  filter(category %in% top_categories)

# Define custom color palette for selected crime categories
custom_colors <- c("red", "pink", "blue", "green", "black")

# Load UK map boundaries for base layer
uk_map <- map_data("world", region = "UK")

# Create an animated plot showing monthly changes in crime hotspots
map_anim_plot <- ggplot() +
  geom_polygon(data = uk_map, aes(x = long, y = lat, group = group),
               fill = "white", color = "gray80") +  # Base map with light outline
  geom_point(data = crime_map_anim, aes(x = long, y = lat, color = category),
             alpha = 0.7, size = 2) +  # Plot crime points with transparency
  scale_color_manual(values = custom_colors) +  # Apply custom colors
  coord_fixed(xlim = range(crime24$long), ylim = range(crime24$lat)) +  # Lock map aspect ratio to data range
  labs(
    title = "Monthly Crime Hotspots in Colchester: {closest_state}",
    subtitle = "Top 5 Crime Categories",
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal() +
  theme(legend.position = "right") +
  transition_states(month, transition_length = 2, state_length = 1) +  # Animate by month
  ease_aes("linear")  # Smooth linear transition

# Render the animation as a GIF
animate(map_anim_plot, nframes = 100, fps = 10, renderer = gifski_renderer())
```

The **animated hotspot map** showcases how the _top five crime categories_ are _spatially distributed_ across **Colchester** each month in **2024**. Each colored point represents an incident from a leading category such as **violent crime**, _anti-social behaviour_, or _shoplifting_.

From this visualization, we observe that _hotspots are consistently concentrated_ around _central Colchester_, with activity fluctuating across months. **Violent crimes** (_black_) appear frequently and are _widely dispersed_, while **shoplifting** (_green_) and _public order offences_ (_blue_) are more centralized. The _dynamic movement and density_ of these hotspots reveal _temporal patterns_ that can aid law enforcement in _allocating resources_ and _planning patrols more effectively_.


### Animated Temperature Over Time
```{r}
# Select relevant columns for animation (daily average temperature)
temp_anim <- temp24 %>%
  select(Date, TemperatureCAvg)

# Create a line plot of daily average temperature
temp_plot <- ggplot(temp_anim, aes(x = Date, y = TemperatureCAvg)) +
  geom_line(color = "purple", size = 0.7) +  # Plot in purple with moderate line thickness
  labs(
    title = "Daily Average Temperature in Colchester (2024)",
    x = "Date", y = "Temperature (°C)"
  ) +
  theme_bw() +  # Use classic black-and-white theme
  transition_reveal(Date)  # Animate reveal over time using the Date variable

# Render the animated plot as a GIF
animate(temp_plot, nframes = 100, fps = 10, renderer = gifski_renderer())
```

The **animated line plot** displays the _daily average temperature_ in **Colchester** throughout **2024**. It clearly reflects the expected _seasonal pattern_—temperatures rise steadily from _winter to summer_, peaking around **July** and **August**, before declining again toward the end of the year.

This smooth fluctuation highlights **Colchester’s temperate climate**, with noticeable _warmth in mid-year_ and _cooler conditions_ during the early and late months. The animation makes it easy to observe _short-term spikes or drops_ in temperature, possibly linked to brief _weather events_.



## **Summary and Recommendations**

### **Summary of Findings**

This project examined whether _weather conditions_—specifically **temperature**, **humidity**, **precipitation**, and **pressure**—influenced _crime patterns_ in **Colchester** during **2024**. Exploratory analysis of both _crime_ and _climate data_ revealed several notable associations.

_Seasonal trends_ showed that crime peaked during **warmer months**, particularly _summer_ and _early autumn_, with offences like _anti-social behaviour_ and _public disorder_ more common during these periods. This suggests that **warmer weather** may indirectly contribute to higher crime through increased _outdoor activity_ and _social interaction_. Conversely, _colder and wetter months_ saw slightly reduced crime, consistent with the idea that _poor weather_ discourages public presence and _opportunistic behaviour_.

Among weather variables, **temperature** had the _strongest association_ with total crime, while _humidity_ and _pressure_ showed weaker, more _category-specific links_. **Rainfall** had minimal effect on overall crime, though certain offences showed _mild sensitivity_ to precipitation.

_Spatial analysis_ identified persistent **hotspots** around _central Colchester_, especially near _nightlife_ and _commercial areas_. **Clustering analysis** revealed _distinct crime zones_, offering useful insights for _targeted resource deployment_. _Animated and interactive visualisations_ further highlighted _temporal_ and _spatial shifts_ in crime.

### **Recommendations**

Based on these findings, several recommendations are proposed. **Policing efforts** could be _adjusted seasonally_, with increased patrols during _warmer months_ and around _hotspot areas_. Integrating **weather data** into _early-warning systems_ may also support **proactive crime prevention**. Additionally, **spatial clustering** results could inform _strategic CCTV placement_ and _community engagement_ initiatives.

### **Future Work**

Future research could build on this project by incorporating additional _environmental factors_ such as **wind speed**, **cloud cover**, and **visibility**, or by examining _crime patterns_ by _time of day_. Applying _predictive modelling techniques_—such as **logistic regression** or **machine learning**—may improve _forecasting of high-risk periods_ based on weather trends. _Comparative studies_ across other towns or cities could also reveal whether similar _weather-crime relationships_ hold in different urban contexts.

### **Conclusion**

This study demonstrates that _weather conditions_ have a _measurable_, though **modest**, impact on _crime patterns_ in **Colchester**. Integrating _environmental insights_ into crime analysis can support more _informed_, _responsive policing_ and long-term **public safety planning**.



## References

- Office for National Statistics (2023). *Crime in England and Wales: year ending December 2022*. Retrieved from https://www.ons.gov.uk

- Essex Police (2024). *Colchester Crime Statistics 2024*. Available via UK Police Data Portal: https://data.police.uk/

- NJ Tierney (2024). *ukp.crime – Street-level crime data documentation*. Retrieved from https://ukpolice.njtierney.com/reference/ukp_crime.html

- B. Czernecki (2024). *meteo_ogimet – Climate data from global weather stations*. Retrieved from https://bczernecki.github.io/climate/reference/meteo_ogimet.html

- Field, A., Miles, J., & Field, Z. (2012). *Discovering Statistics Using R*. SAGE Publications.

- Hipp, J.R., & Kim, Y.A. (2017). “Weather and crime: How weather patterns influence violent and property crimes.” *Environment and Behavior*, 49(2), 143–172.

- Mohler, G.O., Short, M.B., Brantingham, P.J., Schoenberg, F.P., & Tita, G.E. (2011). “Self-exciting point process modeling of crime.” *Journal of the American Statistical Association*, 106(493), 100–108.

- Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York.

- R Core Team (2024). *R: A Language and Environment for Statistical Computing*. R Foundation for Statistical Computing. https://www.R-project.org/

-------------------------------------------------------------------------------------------------------------
